Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-07-10 manuscript_english.Rmd
upgrade manuscirpt english:

We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables. Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.

That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.

This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates. However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.

Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$. Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space. Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.

@@

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$. In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly. in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Es decir:

$$
N_i \propto  S_i(\vec{e}_i)
$$
Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$

@@

1. model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
3. Characterize past analogous climates and retain projections only for sites with climates analogous to the present.
4. Generate a harmonic mean suitability map for all time slices.

@@

### Endemic historic Area
To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $mathbf{S}$ equation. This threshold is the proportion of time that pixel has had ideal conditions. For our study, we consider the value of 90\% since for a period of 120 000 years it is a coverage of 100 000 years, which is a significantly evolutionary time to fix genetic patterns at the population level.

>>>>

For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$. We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$. Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions. If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$. i. e. the set of environmental conditions for which values of fitness are capable of sustaining populations. We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$. This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. and, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$. Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_0(\vec{e}_i) > 1$ and $1$ when $R_0$ is evaluated at the expected value of $\vec {e}(t)$ for all values of the set $\mathbf{N}_F$ where  $R_0(\vec{e}_i) > 1$. In a practical way, instead of $\mathbf{N}_F$, we approximated
$\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F$ 
(the existing niche) from known presence points in the geography $\mathbf{G}$ obtained from emprical observations stored in database, and map to $\mathbf{E}$ in order to numerically fitting a convex volume of the environmental space that enclose the known information.

The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
$$
N_i \propto  S_i(\vec{e}_i)
$$

@@

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and for each trajectory the fitness function changes and therefore the suitability function changes but not in a homogeneous way, so each site presents different displacement rates in envrionmental spaces and as a consequence, different demographic rates. So we can establish that for each site $i$ in geographic space $\mathbf{G}$, there is a population size $N_i$ that changes with respect to a fitness function dependent on time-varying environmental conditions:

$$
N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
$$

@@

1. Model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
3. Characterize past environments and retain projections only for sites with analogous environmental conditions to the present.
4. Calculate the harmonic mean of the suitability for all the sites all time slices.

@@

### Endemic historic Area
To calculate the historic endemic area, the SPA result is taken and a threshold is applied to generate a binary map according to the $\mathbf{S}$ equation. This threshold is the proportion of time that pixel has had ideal conditions. For our study, we consider the value of 90\% since for a period of 120 000 years it is a coverage of 100 000 years, which is a significantly evolutionary time to fix genetic patterns at the population level.

### 2023-07-10 manuscript_english.tex
upgrade manuscirpt english:

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of fitness that depends on environmental
conditions.

That is, the sites in the geography \(\mathbf{G}_o\) that present
positive growth rates are related to favorable conditions. If we take
the set of places in ambient space where these values are greater than a
threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, that
is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

This is a way of expressing the fundamental niche as a function of
environmental conditions. So we can map each point of geography to
ambient space. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive rates. However, we assume that the environment
is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environment and
therefore is different at each time.

Thus locally the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic rate
that depends on the change in environmental conditions over time is:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
any local population will have a positive growth rate.

@@

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
approximated from points of presence in the geography obtained from
databases, numerically approximating a convex surface with a surface
\(\mathbf{N}_R\) of the environmental space.

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and that for each trajectory the fitness function
changes and therefore the suitability function changes but not
accordingly. in a homogeneous way, if not more so that each site
presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of
the function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)).

Es decir:

\[
N_i \propto  S_i(\vec{e}_i)
\] Population size locally is a function of environmental suitability,
where if suitability is \(0\) the population is \(0\) and if suitability
is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\)

@@

To calculate the historic endemic area, the SPA result is taken and a
threshold is applied to generate a binary map according to the
\(mathbf{S}\) equation. This threshold is the proportion of time that
pixel has had ideal conditions. For our study, we consider the value of
90\% since for a period of 120 000 years it is a coverage of 100 000
years, which is a significantly evolutionary time to fix genetic
patterns at the population level.

>>>>

For a given region \(\mathbf{G}\) in te geography, at any point in time,
we have a set of existing environmental conditions
\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the
population's net reproductive rate \(R_0\) is a measure of fitness that
depends on environmental conditions. If we take the set of places in
environmental space where these values are greater than a threshold, we
can define a volume in multidimensional space such that \(R(\vec{e})\)
is always positive and greater than this threshold, that is
\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
environmental conditions for which values of fitness are capable of
sustaining populations. We call
\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
expressing the fundamental niche as a function of environmental
conditions. So we can map each point of geography to ambient space. and,
for each site in the geography there will correspond a site in the
environmental space with different environmental conditions and
different growth rates. However all sites within the region defined by
\(\mathbf{N}_F\) will have viable populations and positive rates
depending on time because we assume that the environment is dynamic and
each geographic point is static, so for each site in geographic space
this growth rate depends on the environment and therefore is different
at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
will be related to the local population size as long as it is within the
boundary defined by \(\mathbf{N}_F\).

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_0(\vec{e}_i) > 1\) and \(1\) when
\(R_0\) is evaluated at the expected value of \(\vec {e}(t)\) for all
values of the set \(\mathbf{N}_F\) where \(R_0(\vec{e}_i) > 1\). In a
practical way, instead of \(\mathbf{N}_F\), we approximated
\(\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F\) (the
existing niche) from known presence points in the geography
\(\mathbf{G}\) obtained from emprical observations stored in database,
and map to \(\mathbf{E}\) in order to numerically fitting a convex
volume of the environmental space that enclose the known information.

The objective of our approach is not to know the function of \(R_0\) but
rather to relate this local variation and the function \(\mathbf{N}_F\)
with a function \$ S\_i(~vec\{e\}\_i)\$ of environmental suitability
that can be estimated from known methods of applied mathematics and
machine learning (for example the case of Maxent, Phillips, Anderson,
and Schapire (2006)).

Population size locally is a function of environmental suitability,
where if suitability is \(0\) the population is \(0\) and if suitability
is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\) \[
N_i \propto  S_i(\vec{e}_i)
\]

@@

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and for each trajectory the fitness function changes
and therefore the suitability function changes but not in a homogeneous
way, so each site presents different displacement rates in envrionmental
spaces and as a consequence, different demographic rates. So we can
establish that for each site \(i\) in geographic space \(\mathbf{G}\),
there is a population size \(N_i\) that changes with respect to a
fitness function dependent on time-varying environmental conditions:

\[
N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
\]

@@

To calculate the historic endemic area, the SPA result is taken and a
threshold is applied to generate a binary map according to the
\(\mathbf{S}\) equation. This threshold is the proportion of time that
pixel has had ideal conditions. For our study, we consider the value of
90\% since for a period of 120 000 years it is a coverage of 100 000
years, which is a significantly evolutionary time to fix genetic
patterns at the population level.
